Versioned data: why it is needed and how it can be achieved (easily and cheaply)

### 2017-11-09 ms.tex
Format for Scientific Data

\documentclass[a4paper,11pt]{article}
\usepackage[osf]{mathpazo}
\usepackage{ms}
\usepackage[]{natbib}
\raggedright

\newcommand{\datastorr}{\texttt{datastorr}}
\newcommand{\smurl}[1]{{\footnotesize\href{https://#1}{#1}}}
\newcommand{\ghsmurl}[1]{{\footnotesize\href{https://github.com/#1}{#1}}}

\usepackage{graphicx}

\title{Versioned data: why it is needed and how it can be achieved (easily and cheaply)}
\author{Daniel S. Falster$^{1,*}$, Richard G. FitzJohn$^2$,\\ Matthew
  W. Pennell$^3$, and William K. Cornwell$^{1}$}
\affiliation{
$^1$ Evolution \& Ecology Research Centre, School of Biological, Earth and Environmental Sciences,
University of New South Wales, Sydney NSW 2052, Australia\\
$^2$ School of Public Health, Imperial College, London SW7 2AZ United Kingdom\\
$^3$ Department of Zoology and Biodiversity Research Centre,\\
University of British Columbia, Vancouver B.C. V6T 1Z4, Canada\\
$^*$ Corresponding author: daniel.falster@unsw.edu.au\\
}
\date{}

\bibliographystyle{mee}

\usepackage[title,titletoc,toc]{appendix}

\mstype{Article Note}
\runninghead{Versioned Data Delivery}
\keywords{Data sharing, Version control, Semantic versioning, Meta-analysis}

\begin{document}
\mstitlepage
\noindent
\parindent=1.5em
\addtolength{\parskip}{.3em}
\doublespacing
\linenumbers

\section{Summary}

The sharing and re-use of data has become a cornerstone of modern science. To help this process, multiple tools allow quick and easy data sharing: a data file can be quickly uploaded, given a digital identifier, and shared. So far, however, data publishing models have not accommodated on-going scientific improvements in data: for many important problems, datasets continue to grow with time -- more data points will be added, and new and more useful data structures will be created. In other words, datasets, like scientific knowledge in general, advance with time. We therefore suggest that many datasets would be more usefully published as a series of versions, with a simple naming system to allow users to perceive the type of change between versions. In this article, we argue that adopting the paradigm and processes for versioned data, analogous to software versioning, would be useful for both data maintainers and users of even small-to-medium sized datasets. We also introduce a system called Versioned Data Delivery (\textsc{vdd}) and present tools for creating, archiving, and distributing versioned data easily, quickly, and cheaply. These new tools allow for individual research groups to shift from a static model of data curation to a dynamic and versioned model that more naturally matches the scientific process.

\section{Introduction}

As is evidenced in the advent of this journal \citep{Editorial-2014}, publication of quality datasets is now considered a first-class scientific product. Increasingly, funding bodies, publishers, and scientific social norms are recognizing the value of sharing data, including as standalone products without any accompanying analyses \citep[e.g.][]{Whitlock-2011,Fairbairn-2011,Piwowar-2011,VanNoorden-2013,Gibney-2013}. Datasets are now routinely archived as part of the publishing process. Moreover, increasing numbers of standalone ``Data papers'' (or descriptors) have been appearing in standard domain-level journals, as well as specialised data journals. Yet, while the last decade has witnessed a rapid and exciting change in attitudes towards data sharing and publishing, the scientific community is still grappling with how to effectively disseminate and manage open-source datasets \citep{Whitlock-2011, Goodman-2014, Lowndes-2017,Perkel-2016,VanNoorden-2013, Kratz-2015}. In particular, the current model for publishing data has not yet embraced the idea that many datasets are designed to answer scientific questions that extend beyond the scope of a single empirical paper. These datasets are constantly evolving, i.e. they are never "finished", and may be the subject of repeated analysis or meta-analysis.

@@

In this article we
\begin{enumerate}
  \item Introduce the concept of Versioned Data Delivery (\textsc{vdd}).
  \item Outline how emerging technologies in data science (Table \ref{tab:technologies}) can be used to help researchers maintain, distribute, and access small-to-medium sized datasets.
  \item Introduce a new \textsc{R} package called \texttt{datastorr}, a proof-of-concept implementation of a \textsc{vdd} system.
\end{enumerate}
The issue of updating and expanding published data has already been addressed in large centralised repositories like genetic sequences (\texttt{GenBank}) or species location data (\texttt{GBIF}), where new data can be added and there exist abilities to correct errors in existing records (e.g. by adding multiple version of a genetic sequence). Yet these ``Dynamic'' web databases (Table \ref{tab:publishing_models}), require a level of infrastructure that is beyond most research groups. Our focus here is on the wide range of datasets, such as those appearing in this journal, that are not covered by these repositories. In most of these cases, the data collected will not be ``big'' but rather small-to-medium sized. Such data support important research projects on particular scientific questions, but are not, in most cases, general enough to warrant custom infrastructure. Further, while we emphasize particular technologies in our implementation, the principles are general and could easily be ported to other platforms.

\section{A lightweight, cheap, and scalable workflow for delivering versioned data}

In brief, the \textsc{vdd} workflow we present borrows best-practices for software development \citep{Perez-Riverol-2016} and applies them to the challenge of maintaining and distributing versioned data. Software developers maintain a core set of code which produces a binary executable file that can be installed on a users local computer. With a \textsc{vdd} system, data developers similarly maintain a core set of files (the ``code''), which produce an organised dataset that can be ``installed'' on a user's local computer. In either the development of software or data, successive versions (called ``releases'') are distributed over time. The similarity in workflow between software and data then allows us to deploy the same technological platforms that are used in software development, for the development and distribution of data. Importantly, our \textsc{vdd} systems uses well-established tools (Table \ref{tab:technologies}), ensuring high-level performance and stability.

@@

\subsection{Version control}

Version control, primarily an open-source variety called \texttt{git}, has become ubiquitous in software development. In practice, version control tracks line-by-line changes in text files and creates and maintains a history of those changes. Increasingly version control has been applied to scientific code and data management, especially for small-to-medium sized datasets \citep{Ram-2013, Perkel-2016, Lowndes-2017}. \texttt{git} is attractive for data management because it tracks all changes in monitored files, provided these are saved in text format (e.g. ``.csv'', ``.tsv'', ``.txt''). The history is visible to anyone interacting with the repository. It also allows users to annotate changes (``commits'') with informative messages detailing the rationale for those changes. In it's present form, \texttt{git} can handle individual data files at least up to 100MB, which includes a large fraction of scientific cases.

@@

\subsection{Hosting and distributing versioned data}

Datasets stored under version control via \texttt{git} reach their real potential when hosted at a suitable internet hosting service \citep{Ram-2013,Perkel-2016}. Here we focus on the platform \texttt{GitHub} (Table \ref{tab:technologies}), but similar functions could be achieved via other providers such as \href{http://bitbucket.org}{bitbucket.org} and \href{http://gitlab.com}{gitlab.com}. Hosting of a \texttt{git} repository enables dataset \emph{maintainers} to connect with other potential \emph{contributors} and also \emph{users} (Fig. \ref{fig:technology_stack}). These platforms are designed to work with \texttt{git} repositories, and thus offer many helpful features, such as ability to record issues, host documentation, or review edits over time.

@@

\subsection{Semantic versioning}

@@

\subsection{\texttt{datastorr} and dataset-specific \texttt{R} packages}

To aid reproducibility and efficient usage, many users will want access to all versions of any particular dataset programmatically (Table \ref{tab:user_requirements}). Code to access a stream of \texttt{GitHub} releases could be written individually by each user, but this creates an unnecessary technological hurdle. To make it easier for users to access versioned data via code, we offer a novel implementation of \textsc{vdd} focussing on the \texttt{R} platform, as one of the most prominent platforms for data science \citep{R-2017}.

Here we introduce a dedicated \textsc{R} package, called \texttt{datastorr} (\smurl{github.com/ropenscilabs/datastorr}), that facilitates access to releases of any versioned data hosted on \texttt{GitHub} (Fig. \ref{fig:technology_stack}). Specifically, the \texttt{datastorr} package:
\begin{enumerate}
  \item Constructs the shell of a second, dataset-specific \textsc{R} package, which is used to access releases from a specific repository stored on \texttt{GitHub};
  \item Contains the main code needed to interact with the \texttt{GitHub} \textsc{api} to retrieve versions of the dataset.
\end{enumerate}
Using \texttt{datastorr}, a researcher can create and distribute a custom \textsc{R} package that facilitates access to their data with (very) minimal computational skills.

For example, \texttt{datastorr} was used to build the package \texttt{baad.data} (\smurl{github.com/traitecoevo/baad.data}), which is an interface to the Biomass and Allometry Database (\textsc{BAAD}) \citep{Falster-2015} stored at \smurl{github.com/dfalster/baad}. The R package \texttt{baad.data} consists of only a few simple functions and associated help files, that were automatically generated with \texttt{datastorr}. For a user, accessing a version of the data is a simple as typing a single line of code (Fig. \ref{fig:technology_stack}). Accessing a different version of the data involves changing only the version number. From the users perspective, the existence of the \texttt{baad.data} and \texttt{datastorr} packages makes reproducing analyses using specific versions of the data possible \citep[e.g.][]{Duursma-2016,Falster-2016}.

@@

\section{Discussion}

@@

A central feature of the proposed  \textsc{vdd} system is that data are maintained in the cloud. This has two main benefits: first, it provides a platform for multiple data contributors to sync their files and correspond about changes in the dataset, and second, it allows for hosting of a stream of data releases for distribution (Fig. \ref{fig:technology_stack}). Cloud systems thus act as a central point for the collection, curation, and distribution of the data. Additionally, one of the greatest benefits of using cloud-based tools like  \texttt{GitHub} for development of software and data has been the way they encourage contributions from multiple individuals working simultaneously --- including from people from outside the initial group of project participants \citep{Rogers-2013, Perkel-2016}. Multiple users can make changes to different parts of the code (or in our case, data) and the \texttt{git} system will integrate these together (if that is possible) or, when needed, flag where there are conflicts that need to be resolved. Adopting a \textsc{vdd} system thus has the added benefit of facilitating seamless and transparent collaboration among research groups in the construction and maintenance of datasets.

@@

The design choices we made in datastorr represent only one of many possible ways to adopt a \textsc{vdd} system; we can imagine many alternative models for interacting with versioned data stored in the cloud. The key is not the specific technology, but rather the concept of creating, maintaining, and distributing versioned data. Indeed, as with every technology the best available approach is certain to evolve, especially as merging technologies --- such as CKAN, OKFN or Git-LFS --- facilitate even better \textsc{vdd} systems in the future.  There are some really exciting ideas that are currently under development which will make versioning systems faster computationally, smaller in file size, flexible with respect to data structures, and thus allow for larger datasets to be versioned efficiently \citep[e.g.,][]{Fli, Dat}.

@@

\section{Acknowledgements}
We thank D Noble for comments on an earlier draft and C Boettiger for helpful discussions. DSF was funded by the Australian Research Council. MWP was funded by a NSERC Discovery Grant.

@@

\section{Tables}

\begin{table}[h!]
\centering
\caption{Alternative frameworks for setting up and managing a living dataset}
{\footnotesize
\vspace{1cm}
  \begin{tabular}{p{2.5cm}p{3.5cm}p{3.5cm}p{4cm}}
  \hline
  \textbf{Feature} & \textbf{Static datasets}& \textbf{Web database} & \textbf{Versioned Data Delivery}\\
  \textbf{} & (e.g. \smurl{datadryad.org})& (e.g. \smurl{coraltraits.org}) & (eg. \texttt{datastorr} via \texttt{GitHub})\\
  \hline
   Webserver        & bespoke & bespoke &  \smurl{github.com}\\
   Backend          & none & SQL + Ruby-on-rails 			& \texttt{git} + \texttt{datastorr} \\
   User access      & Web browser & Web browser 				    & Web browser or R \\
   Ease of setup    & Very easy & Hard 							& Easy\\
   Data size        & Up to several Gb & Small-Very large 				& Up to 1Gb\\
   Cost             & Varies & Varies  						& Free \\
   Bandwidth        & Managed by provider & Pro rata 						& Managed by \texttt{GitHub}\\
   Maintainer skills & None & Ruby + php 					& R + \texttt{git} \\
   User skills      &Web browsing& Web browsing  					& basic R \\
   Versioning       &None& Hard 							& Easy \\
   \textsc{doi} minting      &Automatic & Manual 					& Automatic \\
  \hline 
  \\

@@

\begin{table}[h!]
\centering
\caption{Overview of technologies used to maintain, store, and distribute the versioned data as described in this paper.}
{\footnotesize
\vspace{1cm}
  \begin{tabular}{p{3cm}p{12cm}}
  \hline
  \textbf{Technology} & \textbf{Description} \\\hline
  \textsc{API} & An Application Programming Interface provides a set of protocols for exchanging information.\\
   \texttt{datastorr} & R package used to fetch versioned releases from \texttt{GitHub}. \\
   \textsc{doi} & Digital object identifier, which refers a user to a single published object. \\
   \texttt{git} & Open source version control system used for tracking progressive changes in a set of text files, typically computer code but also data.\\
   github.com & A commercial web platform for sharing, visualising, and managing `git' repositories. Includes ability to browse the `history', `issue' tracking, and ability to create `releases'.\\
   \texttt{R} & Open source statistical and data processing language.\\
   \hline
  \end{tabular}
  }
\label{tab:technologies}
\end{table}

@@

\begin{table}[h!]
\centering
\caption{Groups of users interacting with the Versioned Data Delivery system described and their requirements.}
{\footnotesize
\vspace{1cm}
  \begin{tabular}{p{2cm}p{5cm}p{7cm}}
  \hline
% Latex being the worst has made the overflow in the second column flow through to affect the third. I used to know how to deal with this...
  \textbf{Group} & \textbf{Goal} & \textbf{Requirements} \\ \hline
  Maintainer & Create and distribute versioned datasets & Low technical overhead \\
    & & Easy workflow for releasing new versions \\
    & & Long term preservation \\
    & & Easy to crowd-source error checking and contributions \\
    & & Low initial cost \\
    & & Low on-going maintenance \\
  Contributor & Contribute to future versions of a dataset & Add new data \\
    & & Report errors in existing data \\
  Users (all) & Easy access to releases from a versioned dataset & Introduction \& overview \\
    & & Long term stability \\
    & & Clear path for users to become contributors \\
  Users (programmatic) & Build reproducible products using specific versions of a dataset & Programmatic access to releases \\
    & & Easy installation \\
    & & Long term stability \\
  \hline
  \end{tabular}
}
\label{tab:user_requirements}
\end{table}

@@

  \begin{tabular}{p{3.5cm}p{3cm}p{7cm}}
  \hline
   \textbf{\texttt{GitHub} repo} & \textbf{\texttt{R} package} & \textbf{Description} \\ \hline
  \ghsmurl{dfalster/baad} & \texttt{baad.data} & The \texttt{Biomass And Allometry Database} provides data on the size dimensions of plants for many species, compiled from multiple scientific papers \citep{Falster-2015}.\\
  \ghsmurl{traitecoevo/taxonlookup} & \texttt{taxonlookup} & Provides a taxonomic lookup table for land plants \citep{Pennell-2015a}.\\
  \hline
  \end{tabular}
  }
\label{tab:examples}
\end{table}

@@

\section{Figures}

>>>>

\documentclass[english]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{babel}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{fancyhdr}
\usepackage[colorlinks=false, allcolors=grey, breaklinks=true]{hyperref}

\RequirePackage[superscript,biblabel,nomove]{cite}
\bibliographystyle{assets/naturemag-doi}
\RequirePackage{assets/jabbrv}

%\bibliographystyle{naturemag}
\newcommand{\scidatalogo}{\includegraphics[height=36pt]{assets/SciData_logo.jpg}}
\pagestyle{fancy}
\fancyhf{}
\renewcommand{\headrulewidth}{0pt}
\setlength{\headheight}{40pt} 
\lhead{\textsc{\scidatalogo}}
\rhead{Falster \emph{et al.} -- Versioned Data Delivery}

\newcommand{\datastorr}{\texttt{datastorr}}
\newcommand{\smurl}[1]{{\footnotesize\href{https://#1}{\nolinkurl{#1}}}}
\newcommand{\ghsmurl}[1]{{\footnotesize\href{https://github.com/#1}{\nolinkurl{#1}}}}

\begin{document}

\title{Versioned data: why it is needed and how it can be achieved (easily and cheaply)}

\author{Daniel S. Falster\textsuperscript{1,*}, Richard G. FitzJohn\textsuperscript{2},\\ Matthew
  W. Pennell\textsuperscript{3}, and William K. Cornwell\textsuperscript{1}}

\maketitle
\thispagestyle{fancy}
$^1$ Evolution \& Ecology Research Centre, School of Biological, Earth and Environmental Sciences,
University of New South Wales, Sydney NSW 2052, Australia; 
$^2$ School of Public Health, Imperial College, London SW7 2AZ United Kingdom; 
$^3$ Department of Zoology and Biodiversity Research Centre, University of British Columbia, Vancouver B.C. V6T 1Z4, Canada;
$^*$ Corresponding author: Daniel Falster (daniel.falster@unsw.edu.au)

\begin{abstract}
The sharing and re-use of data has become a cornerstone of modern science. Multiple platforms now allow quick and easy data sharing. So far, however, data publishing models have not accommodated on-going scientific improvements in data: for many problems, datasets continue to grow with time -- more records are added, errors fixed, and new data structures are created. In other words, datasets, like scientific knowledge, advance with time. We therefore suggest that many datasets would be usefully published as a series of versions, with a simple naming system to allow users to perceive the type of change between versions. In this article, we argue for adopting the paradigm and processes for versioned data, analogous to software versioning. We also introduce a system called Versioned Data Delivery and present tools for creating, archiving, and distributing versioned data easily, quickly, and cheaply. These new tools allow for individual research groups to shift from a static model of data curation to a dynamic and versioned model that more naturally matches the scientific process.
\end{abstract}

%\keywords{Data sharing, Version control, Semantic versioning, Meta-analysis}

\section*{Background}

As is evidenced in the advent of this journal \cite{Editorial-2014}, publication of quality datasets is now considered a first-class scientific product. Increasingly, funding bodies, publishers, and scientific social norms are recognizing the value of sharing data, including as standalone products without any accompanying analyses \cite{Whitlock-2011,Fairbairn-2011,Piwowar-2011,VanNoorden-2013,Gibney-2013}. Datasets are now routinely archived as part of the publishing process. Moreover, increasing numbers of standalone ``Data papers'' (or descriptors) have been appearing in standard domain-level journals, as well as specialised data journals. Yet, while the last decade has witnessed a rapid and exciting change in attitudes towards data sharing and publishing, the scientific community is still grappling with how to effectively disseminate and manage open-source datasets \cite{Whitlock-2011, Goodman-2014, Lowndes-2017,Perkel-2016,VanNoorden-2013, Kratz-2015}. In particular, the current model for publishing data has not yet embraced the idea that many datasets are designed to answer scientific questions that extend beyond the scope of a single empirical paper. These datasets are constantly evolving, i.e. they are never ``finished'', and may be the subject of repeated analysis or meta-analysis.

@@

In this article we 1) Introduce the concept of Versioned Data Delivery (\textsc{vdd}); 2) Outline how emerging technologies in data science (Table \ref{tab:technologies}) can be used to help researchers maintain, distribute, and access small-to-medium sized datasets; and 3) Introduce a new \textsc{R} package called \texttt{datastorr}, a proof-of-concept implementation of a \textsc{vdd} system. The issue of updating and expanding published data has already been addressed in large centralised repositories like genetic sequences (\texttt{GenBank}) or species location data, where new data can be added and there exist abilities to correct errors in existing records (e.g. by adding multiple version of a genetic sequence). Yet these ``Dynamic'' web databases (Table \ref{tab:publishing_models}), require a level of infrastructure that is beyond most research groups. Our focus here is on the wide range of datasets, such as those appearing in this journal, that are not covered by these repositories. In most of these cases, the data collected will not be ``big'' but rather small-to-medium sized. Such data support important research projects on particular scientific questions, but are not, in most cases, general enough to warrant custom infrastructure. Further, while we emphasize particular technologies in our implementation, the principles are general and could easily be ported to other platforms.

\section*{A lightweight, cheap, and scalable workflow for delivering versioned data}

In brief, the \textsc{vdd} workflow we present borrows best-practices for software development \cite{Perez-Riverol-2016} and applies them to the challenge of maintaining and distributing versioned data. Software developers maintain a core set of code which produces a binary executable file that can be installed on a users local computer. With a \textsc{vdd} system, data developers similarly maintain a core set of files (the ``code''), which produce an organised dataset that can be ``installed'' on a user's local computer. In either the development of software or data, successive versions (called ``releases'') are distributed over time. The similarity in workflow between software and data then allows us to deploy the same technological platforms that are used in software development, for the development and distribution of data. Importantly, our \textsc{vdd} systems uses well-established tools (Table \ref{tab:technologies}), ensuring high-level performance and stability.

@@

\subsection*{Version control}

Version control, primarily an open-source variety called \texttt{git}, has become ubiquitous in software development. In practice, version control tracks line-by-line changes in text files and creates and maintains a history of those changes. Increasingly version control has been applied to scientific code and data management, especially for small-to-medium sized datasets \cite{Ram-2013, Perkel-2016, Lowndes-2017}. \texttt{git} is attractive for data management because it tracks all changes in monitored files, provided these are saved in text format (e.g. ``.csv'', ``.tsv'', ``.txt''). The history is visible to anyone interacting with the repository. It also allows users to annotate changes (``commits'') with informative messages detailing the rationale for those changes. In it's present form, \texttt{git} can handle individual data files at least up to 100MB, which includes a large fraction of scientific cases.

@@

\subsection*{Hosting and distributing versioned data}

Datasets stored under version control via \texttt{git} reach their real potential when hosted at a suitable internet hosting service \cite{Ram-2013,Perkel-2016}. Here we focus on the platform \texttt{GitHub} (Table \ref{tab:technologies}), but similar functions could be achieved via other providers such as \href{http://bitbucket.org}{bitbucket.org} and \href{http://gitlab.com}{gitlab.com}. Hosting of a \texttt{git} repository enables dataset \emph{maintainers} to connect with other potential \emph{contributors} and also \emph{users} (Fig. \ref{fig:technology_stack}). These platforms are designed to work with \texttt{git} repositories, and thus offer many helpful features, such as ability to record issues, host documentation, or review edits over time.

@@

\subsection*{Semantic versioning}

@@

\subsection*{\texttt{datastorr} and dataset-specific \texttt{R} packages}

To aid reproducibility and efficient usage, many users will want access to all versions of any particular dataset programmatically (Table \ref{tab:user_requirements}). Code to access a stream of \texttt{GitHub} releases could be written individually by each user, but this creates an unnecessary technological hurdle. To make it easier for users to access versioned data via code, we offer a novel implementation of \textsc{vdd} focussing on the \texttt{R} platform, as one of the most prominent platforms for data science \cite{R-2017}.

Here we introduce a dedicated \textsc{R} package, called \texttt{datastorr} (\smurl{github.com/ropenscilabs/datastorr}), that facilitates access to releases of any versioned data hosted on \texttt{GitHub} (Fig. \ref{fig:technology_stack}). Specifically, the \texttt{datastorr} package: 1)Constructs the shell of a second, dataset-specific \textsc{R} package, which is used to access releases from a specific repository stored on \texttt{GitHub}; and 2) Contains the main code needed to interact with the \texttt{GitHub} \textsc{api} to retrieve versions of the dataset. 
Using \texttt{datastorr}, a researcher can create and distribute a custom \textsc{R} package that facilitates access to their data with (very) minimal computational skills.

For example, \texttt{datastorr} has been used to build several packages (Table \ref{tab:examples}), including \texttt{baad.data} (\smurl{github.com/traitecoevo/baad.data}), which is an interface to the Biomass and Allometry Database \cite{Falster-2015} stored at \smurl{github.com/dfalster/baad}. The R package \texttt{baad.data} consists of only a few simple functions and associated help files, that were automatically generated with \texttt{datastorr}. For a user, accessing a version of the data is a simple as typing a single line of code (Fig. \ref{fig:technology_stack}). Accessing a different version of the data involves changing only the version number. From the users perspective, the existence of the \texttt{baad.data} and \texttt{datastorr} packages makes reproducing analyses using specific versions of the data possible \cite{Duursma-2016,Falster-2016}.

@@

\section*{Discussion}

@@

A central feature of the proposed  \textsc{vdd} system is that data are maintained in the cloud. This has two main benefits: first, it provides a platform for multiple data contributors to sync their files and correspond about changes in the dataset, and second, it allows for hosting of a stream of data releases for distribution (Fig. \ref{fig:technology_stack}). Cloud systems thus act as a central point for the collection, curation, and distribution of the data. Additionally, one of the greatest benefits of using cloud-based tools like  \texttt{GitHub} for development of software and data has been the way they encourage contributions from multiple individuals working simultaneously --- including from people from outside the initial group of project participants \cite{Rogers-2013, Perkel-2016}. Multiple users can make changes to different parts of the code (or in our case, data) and the \texttt{git} system will integrate these together (if that is possible) or, when needed, flag where there are conflicts that need to be resolved. Adopting a \textsc{vdd} system thus has the added benefit of facilitating seamless and transparent collaboration among research groups in the construction and maintenance of datasets.

@@

The design choices we made in datastorr represent only one of many possible ways to adopt a \textsc{vdd} system; we can imagine many alternative models for interacting with versioned data stored in the cloud. The key is not the specific technology, but rather the concept of creating, maintaining, and distributing versioned data. Indeed, as with every technology the best available approach is certain to evolve, especially as merging technologies --- such as CKAN, OKFN or Git-LFS --- facilitate even better \textsc{vdd} systems in the future.  There are some really exciting ideas that are currently under development which will make versioning systems faster computationally, smaller in file size, flexible with respect to data structures, and thus allow for larger datasets to be versioned efficiently \cite{Fli, Dat}.

@@

\section*{Acknowledgements}
We thank D Noble for comments on an earlier draft and C Boettiger for helpful discussions. DSF was funded by the Australian Research Council. MWP was funded by a NSERC Discovery Grant.

RF developed the datastorr package. All authors discussed the concepts and wrote the paper.

@@

\section*{Tables}

\begin{table}[h!]
\centering
\caption{Alternative frameworks for distributing data.}
{\footnotesize
\vspace{1cm}
  \begin{tabular}{p{2.5cm}p{3cm}p{3cm}p{4cm}}
  \hline
  \textbf{Feature} & \textbf{Static datasets}& \textbf{Web databases} & \textbf{Versioned Data Delivery}\\
   Platform (e.g.) 	& datadryad.org 			& coraltraits.org	  &  github.com\\
   User access      & Web browser 				& Web browser 			& Web browser or R \\
   Ease of setup    & Very easy 					& Hard 							& Easy\\
   Data size        & Up to several Gb 		& Small-Very large 	& Up to 1Gb\\
   Cost             & Varies 							& Varies  					& Free \\
   Bandwidth        & Managed by provider & Pro rata 					& Managed by \texttt{GitHub}\\
   Maintainer skills & None 							& High							& R + \texttt{git} \\
   User skills      & Web browsing 	& Web browsing  					& Web browsing or R \\
   Versioning       & None 					& Hard 										& Easy \\
   \textsc{doi} minting  &Automatic & Manual 									& Automatic \\
  \hline 
  \\

@@

\begin{table}[h!]
\centering
\caption{Overview of technologies used to maintain, store, and distribute the versioned data as described in this paper.}
{\footnotesize
\vspace{1cm}
  \begin{tabular}{p{3cm}p{9cm}}
  \hline
  \textbf{Technology} & \textbf{Description} \\\hline
  \textsc{API} & An Application Programming Interface provides a set of protocols for exchanging information.\\
   \texttt{datastorr} & R package used to fetch versioned releases from \texttt{GitHub}. \\
   \textsc{doi} & Digital object identifier, which refers a user to a single published object. \\
   \texttt{git} & Open source version control system used for tracking progressive changes in a set of text files, typically computer code but also data.\\
   github.com & A commercial web platform for sharing, visualising, and managing `git' repositories. Includes ability to browse the `history', `issue' tracking, and ability to create `releases'.\\
   \texttt{R} & Open source statistical and data processing language.\\
   \hline
  \end{tabular}
  }
\label{tab:technologies}
\end{table}

@@

\begin{table}[h!]
\centering
\caption{Groups of users interacting with the Versioned Data Delivery system described and their requirements.}
{\footnotesize
\vspace{1cm}
  \begin{tabular}{p{2cm}p{4cm}p{6cm}}
  \hline
% Latex being the worst has made the overflow in the second column flow through to affect the third. I used to know how to deal with this...
  \textbf{Group} & \textbf{Goal} & \textbf{Requirements} \\ \hline
  Maintainer & Create and distribute versioned datasets & Low technical overhead \\
    & & Easy workflow for releasing new versions \\
    & & Long term preservation \\
    & & Easy to crowd-source error checking and contributions \\
    & & Low initial cost \\
    & & Low on-going maintenance \\
  Contributor & Contribute to future versions of a dataset & Add new data \\
    & & Report errors in existing data \\
  Users (all) & Easy access to releases from a versioned dataset & Introduction \& overview \\
    & & Long term stability \\
    & & Clear path for users to become contributors \\
  Users (programmatic) & Build reproducible products using specific versions of a dataset & Programmatic access to releases \\
    & & Easy installation \\
    & & Long term stability \\
  \hline
  \end{tabular}
}
\label{tab:user_requirements}
\end{table}

@@

  \begin{tabular}{p{3.5cm}p{2cm}p{6.5cm}}
  \hline
   \textbf{\texttt{GitHub} repo} & \textbf{\texttt{R} package} & \textbf{Description} \\ \hline
  \ghsmurl{dfalster/baad} & \texttt{baad.data} & The \texttt{Biomass And Allometry Database} provides data on the size dimensions of plants for many species, compiled from multiple scientific papers \cite{Falster-2015}.\\
  \ghsmurl{traitecoevo/taxonlookup} & \texttt{taxonlookup} & Provides a taxonomic lookup table for land plants \cite{Pennell-2015a}.\\
  \hline
  \end{tabular}
  }
\label{tab:examples}
\end{table}

@@

\section*{Figures}
